\large{Vipassana for Hackers} \\ \Huge{Paper Two: The Brain} \\ \large\textit{Version 0.2}

### 2019-10-09 paper-two/paper-two.tex
updated info on consciousness

This is hard at first, with the body involuntarily reacting by jerking and writhing
in involuntary response to the very act of observation. Initially, even awareness
itself has a jerky quality and it is hard to observe sensations consciously. There is also the
pain. This eclipses all other sensations. In my case, it was difficult to maintain
awareness of any other sensation with the pain that was emanating from my legs and
back. The OFC is involved in assigning emotion to sensation \cite{ofcemotion} and the
ACC plays a vital role in perception of sensation. These two parts of the brain are
usually functionally connected. In the beginning, as I sat trying to observe
sensations and was only aware of the pain, the ACC was firing rapidly and the
OFC then assigned these signals to the Limbic System, which
put me in a less than ideal emotional state.

@@

Perhaps the strangest experience during a Vipassana course is that of shifts in the
sense of self. The ``self'', as a subjective agent in space and time, changes. At
first, this change is only very slight. As a student closes her eyes and begins to
meditate, it may feel as though the body has shifted in space or rotated in different direction, 
due to the temporary down regulation of the prefrontal cortex (don't worry about it, keep observing
sensations). \todo{So? What does this mean?} As the meditations get deeper the DMN
occasionally quiets completely (perhaps only for a short duration) while the AIC
fires more rapidly, leading to alternative networks of consciousness
arising. \todo{What does this mean? What is an example of ``alternative networks of consciousness''?} It is
as if losing the influence of one's own narrative leads to new insights that are
usually kept from consciousness. It has been found that experiencing this, even for a
few seconds, can lead to lasting changes long after the activity of meditation itself
has stopped. \cite{alteredtraits} In the moment, these experiences can be quite
similar to being on certain hallucinogenic drugs or even, in some cases, to ecstatic
seizures where the abnormal activity of the Anterior Insular Cortex leads to
heightened self-awareness, feelings of bliss, and a lack of ambiguity. \cite{cortexbliss}

@@

>>>>

This is hard at first, with the body involuntarily reacting by jerking and writhing
in involuntary response to the very act of observation. Initially, even awareness
itself has a jerky quality and it is hard to observe sensations consciously. There is also the
pain. This eclipses all other sensations. In my case, it was difficult to maintain
awareness of any other sensation with the pain that was emanating from my legs and
back. The OFC is involved in assigning emotion to sensation and the shaping of pain by expectations
\cite{ofcemotion,ofcexpect} and the ACC plays a vital role in perception of sensation. 
These two parts of the brain are usually functionally connected. 
In the beginning, as I sat trying to observe sensations and was only aware of the pain, 
the ACC was firing rapidly and the OFC then assigned these signals to the Limbic System, which
put me in a less than ideal emotional state.

@@

Perhaps the strangest experience during a Vipassana course is that of shifts in the
sense of self. The ``self'', as a subjective agent in space and time, changes. At
first, this change is only very slight. As a student closes her eyes and begins to
meditate, it may feel as though the body has shifted in space or rotated in different direction, this is
due to the temporary down regulation of the prefrontal cortex.
 \todo{So? What does this mean?} As the meditations get deeper the DMN
occasionally quiets completely (perhaps only for a short duration) while the AIC
fires more rapidly, leading to alternative networks of consciousness
arising. \todo
Consciousness is defined as our subjective awareness of ourselves and our environment. 
An altered state of consciousness is defined as alternate patterns or configurations of 
this experience, which differ qualitatively from a baseline state.
Altered states of consciousness

{What does this mean? What is an example of ``alternative networks of consciousness''?} 
It is as if losing the influence of one's own narrative leads to new insights that are
usually kept from consciousness. It has been found that experiencing this, even for a
few seconds, can lead to lasting changes long after the activity of meditation itself
has stopped. \cite{alteredtraits} In the moment, these experiences can be quite
similar to being on certain hallucinogenic drugs or even, in some cases, to ecstatic
seizures where the abnormal activity of the Anterior Insular Cortex leads to
heightened self-awareness, feelings of bliss, and a lack of ambiguity. \cite{cortexbliss}

@@

\bibitem{ofcexpect}
 Lauren Y.Atlas, Tor D.Wager
 \textit{How expectations shape pain.}
  Neuroscience Letters,  520 (2):140-148, 29 June 2012.
